A formal treatment of bidirectional typing

### 2023-10-15 tex/ESOP24/S6-example.tex
[ tex ] Update Future work

\section{Examples}\label{sec:example}
Now we use two more examples to exhibit the applicability of our approach.
This first one has infinitely many operations and the second one includes many more constructs than simply typed $\lambda$-calculus with some common features in a programming language to justify the practical needs for a general development.

\subsection{Spine application}\label{subsec:spine}
A spine application $t\;u_1\;\ldots\;u_n$ is a variant form of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
This arrangement allows direct access to the head term, making it practical in various applications.
For instance, \Agda's core language employs this form of application, as does its reflected syntax used for metaprogramming.

At first glance, accommodating this form of application may seem impossible, given that the number of arguments for a construct is finite and has to be fixed.
Nonetheless, there is no constraint on the total number of operation symbols a signature can have, allowing us to establish a corresponding construct for each number $n$ of arguments, i.e.\ viewing the following rule
\bgroup
\small
  \begin{mathpar}
    \inferrule{\Gamma \vdash t \syn A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right) \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
  \end{mathpar}
\egroup
as a rule schema parametrised by $n$.
Then, we use the same type signature~$\Sigma_{\bto}$ but extend the signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
\[
  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right)^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
\]
so each application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$, thereby exhibiting the necessity of having an arbitrary set for operation symbols.

\subsection{Computational calculi}\label{subsec:PCF}
The code size of a type synthesiser grows proportionally to the number of type constructs and the number of arguments for each language construct, making it a tedious task to implement.
Even for a simply typed language which is generally straightforward to implement a synthesiser individually, having a generator can still reduce the effort significantly.
For illustrative purposes, let us consider the Moggi's computational $\lambda$-calculus~\cite{Moggi1989} with additional constructs.

We first introduce additional type constructs apart from $\bto$ by extending $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
\[
  \arity({\mathsf{nat}}) = 0\qquad
  \arity({\mathsf{T}}) = 1\qquad
  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2
\]
introducing $\mathtt{nat}$, $A \times B$, $A + B$, $T(A)$ as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
The bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations in following table.

%\begin{table}
\bgroup
  \renewcommand{\arraystretch}{2}
  \centering\small
\begin{tabular}{c | l}
  Rules & Operations \\ \hline\hline
  $\inferrule{}{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}$ &  $\aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}$ 
  \\
 $\inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}$ & $\aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}$ & $\aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}$ & $\aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}$  \\
 $\inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}$ & $\aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}$ \\

 $\inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}$ & $\aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\chk}}$ \\

 $\inferrule{\Gamma \vdash t \chk A}{\Gamma \vdash \mathtt{inj}_1(t) \chk A + B}$ & $\aritysymbol{\mathsf{inj}_1}{A, B}{A^{\chk}}{A + B^{\chk}}$ \\

 $\inferrule{\Gamma \vdash u \chk B}{\Gamma \vdash \mathtt{inj}_2(u) \chk A + B}$ & $\aritysymbol{\mathsf{inj}_2}{A, B}{B^{\chk}}{A + B^{\chk}}$ \\

@@

 $\inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}$ & 
  $\aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \syn A}{\Gamma \vdash \mathtt{ret}(t) \syn T(A)}$ & $\aritysymbol{\mathsf{ret}}{A}{A^{\syn}}{T(A)^{\syn}}$ \\
 $\inferrule{\Gamma \vdash t \syn T(A) \and \Gamma, x : A \vdash u \syn T(B)}{\Gamma \vdash \mathtt{bind}(t; x.u) \syn T(B)}$ & $\aritysymbol{\mathsf{bind}}{A, B}{T(A)^{\syn}, [A]T(B)^{\syn}}{T(B)^{\syn}}$
\end{tabular}
\egroup

The extended language has `only' 15 constructs, far fewer than what the surface language of a programming language in use would have. 
However, if we implement a proof-relevant type synthesiser in the style of \citet{Wadler2022}, then there will be nearly 100 cases to check on top of other auxiliary definitions.
On the other hand, similar to a parser generator, a type-synthesiser generator only needs a language specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
We will see in the next section how the actual formalisation is used.

>>>>

\section{Examples}\label{sec:example}
To exhibit the applicability of our approach, we discuss two more examples:
one has infinitely many operations and the other includes many more constructs than simply typed $\lambda$-calculus, exhibiting the practical side of a general treatment.

\subsection{Spine application}\label{subsec:spine}
A spine application $t\;u_1\;\ldots\;u_n$ is a form of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
This arrangement allows direct access to the head term, making it practical in various applications, and has been used by \Agda's core language. %employs this form of application, as does its reflected syntax used for metaprogramming.

At first glance, accommodating this form of application may seem impossible, given that the number of arguments for a construct is finite and has to be fixed.
Nonetheless, the total number of operation symbols a signature can have need not be finite, allowing us to establish a corresponding construct for each number $n$ of arguments, i.e.\ viewing the following rule
\bgroup
\small
  \begin{mathpar}
    \inferrule{\Gamma \vdash t \syn A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right) \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
  \end{mathpar}
\egroup
as a rule schema parametrised by $n$, so the signature $\Omega_{\Lambda}^{\Leftrightarrow}$ can be extended with 
\[
  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right)^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
\]
Each application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$, thereby exhibiting the necessity of having an arbitrary set for operation symbols.

\subsection{Computational calculi}\label{subsec:PCF}
Implementing a standalone type synthesizer for a simply typed language is typically a straightforward task.
However, the code size increases proportionally to the number of type constructs and of arguments associated with each term construct.
When dealing with a fixed number of $n$ type constructs, for each synthesising construct, there are two cases for a checking argument but $n + 1$ possible cases for each synthesising argument---the successful synthesis of the expected type, instances where it fails, or $n-1$ cases where the expected type does not match; similarly for a checking construct---making the task tedious and error-prone.
Therefore, having a generator can be helpful and significantly reduce the effort for implementation.

For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with additional constructs listed in \cref{tab:computational-calculus}.
The extended language has `only' 15 constructs, far fewer than a realistic programming language would have, but there are nearly 100 possible cases already to consider during bidirectional type synthesis.
\begin{table}[h!]
  \renewcommand{\arraystretch}{2.5}
  \setlength{\tabcolsep}{3pt}
  \centering\footnotesize
\begin{tabular}{c | l}
  Rules & Operations \\ \hline\hline
  $\inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}$ &  $\aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}$ 
  \\
 $\inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}$ & $\aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}$ & $\aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}$ & $\aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}$  \\
 $\inferrule{\Gamma \vdash t \syn A_1 \times A_2}{\Gamma \vdash \mathtt{proj}_i(t) \syn A_i}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{proj}_i}{A_1, A_2}{A_1 \times A_2^{\syn}}{A_i^{\syn}}$ for $i = 1, 2$ \\

 $\inferrule{\Gamma \vdash t \chk A_i}{\Gamma \vdash \mathtt{inj}_i(t) \chk A_1 + A_2}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{inj}_i}{A_1, A_2}{A_i^{\chk}}{A_1 + A_2^{\chk}}$ for $i = 1, 2$ \\

@@

 $\inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}$ & 
  $\aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \syn A}{\Gamma \vdash \mathtt{ret}(t) \syn T(A)}$ & $\aritysymbol{\mathsf{ret}}{A}{A^{\syn}}{T(A)^{\syn}}$ \\
 $\inferrule{\Gamma \vdash t \syn T(A) \and \Gamma, x : A \vdash u \syn T(B)}{\Gamma \vdash \mathtt{bind}(t; x.u) \syn T(B)}$ & $\aritysymbol{\mathsf{bind}}{A, B}{T(A)^{\syn}, [A]T(B)^{\syn}}{T(B)^{\syn}}$
\end{tabular}
\caption{A computational calculus with naturals, products, sums, and general recursion}
\label{tab:computational-calculus}
\end{table}

On the other hand, similar to a parser generator, a type-synthesiser generator only needs a specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
In more detail, such a specification can be derived by extending $\Sigma_{\bto}$ with additional type constructs with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
\[
  \arity({\mathsf{nat}}) = 0\qquad
  \arity({\mathsf{T}}) = 1\qquad
  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2.
\]
Types $\mathtt{nat}$, $A \times B$, $A + B$, and $T(A)$ are introduced as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
Finally, invoking \cref{cor:trichotomy} formally with the specification~$(\Sigma, \Omega)$ gives us a type synthesiser directly.\footnote{%
See \cref{sec:formalisation} for a demonstration in \Agda.}

### 2023-10-15 tex/ESOP24/S8-future.tex
[ update ] Add the missing .bib
[ tex ] Update Future work

\section{Discussion and Future work} \label{sec:future}

\subsection{Related work}
\citet{Harper1993a,Harper2007}

Our general definitions are akin to the definitions used in other language formalisation frameworks~\citep{Ahrens2022,Allais2021,Fiore2022} in a proof assistant---the concept of a signature is at object level (i.e.\ an inhabitant of a record type in \Agda).
This allows us to exploit the expressive power of our meta-language while defining a language, so it is possible to express a construct that takes an indeterminate number of arguments such as spine application.
%\todo[inline,caption={}]{\begin{enumerate}
%\item `Input' and `output' of our theory, comparing with grammars and parsing
%\item A more comprehensive theory than Wadler et al., and more formal and refined than Dunfield et al.'
%\item Restrictions to simple types (not individual language features) and syntax-directedness
%\end{enumerate}}

%which together describe an \emph{extrinsically typed} and \emph{syntax-directed} bidirectional simple type system~(\cref{fig:raw-terms,fig:bidirectional-typing-derivations}).
%
%There are many ways and extensions for designing bidirectional typing rules, .
%Yet, once the definition of a bidirectional type system is in place, deriving an algorithm becomes straightforward.
%Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce a formalism for bidirectional typing.
%\todo{Looks like we should have axiomatised what types we can deal with; left as future work}
\paragraph{More characteristics of bidirectional typing}

We have formalised some of the most important concepts discussed by \citet{Dunfield2021}.
There are other concepts that may be formalisable with some more effort, for example Pfenning's recipe for bidirectionalising typing rules.
We believe that our theory lays the foundation for further formalising these concepts.
There are some other concepts that may be more difficult to pin down, notably `annotation character', which is roughly about how easy it is for the user to write annotated programs.
These may continue to require creativity from the designer of type systems.

%\paragraph{Beyond syntax-directed type systems}
%
%As a first theory of bidirectional type synthesis, our work focuses on syntax-directed type systems.
%To explore more general settings, an extreme possibility is to independently specify an ordinary type system and a bidirectional one over the same raw terms and then investigate possible relationships between the two systems, but a more manageable and practical setting suitable for a next step is probably one that only mildly generalises ours: 
%for each raw term construct, there can be several ordinary typing rules, and each typing rule can be refined to several bidirectional typing rules with different mode assignments.
%Even for this mildly generalised setting, there is already a lot more work to do:
%Both the mode decorator and the bidirectional type synthesiser will have to backtrack, and become significantly more complex.
%For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, but the completeness direction will pose a problem---for every node of a raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
%There should be ways to fix this problem while retaining the mode decoration phase---for example, one idea is to make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of these mode derivations into a bidirectional typing derivation.
%All these considerations are rather too complex for a first theory, however.
%
%%\Josh[inline, caption={}]{Getting rid of the assumption of being syntax-directed (the triangular picture) --- 0.5p
%%\begin{enumerate}
%%\item In general: two independently formulated type systems over raw terms
%%\item More sensible next step: a `refinement chain'
%%\item Backtracking algorithms are required. for both mode decoration and bidirectional type synthesis.
%%\item The idea of separating and combining mode and type information should still work, but requires more care in the completeness direction.
%%\item Too complex for a first theory, and left as future work
%%\end{enumerate}}
%
\paragraph{Beyond simple types}
We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}).
While algebraic approaches to polymorphic types have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
Subtyping is essential for formulating important concepts such as \emph{principal types} in type synthesis.

%%\paragraph{Whither extrinsic typing?}
%
%%\todo[inline]{We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}), which requires advancement in language formalisation~(\cref{sec:language-formalisation}) that is orthogonal to our work.}
%
%%\Josh[inline]{Optional: ornaments (0.25p), NDGP (0.25p)}
%
%%\subsubsection{Theories of abstract syntax with variable binding}
%%\label{sec:theory-of-syntax}
%%
%
%%Substitution is also modelled categorically, but it does not play a role in this paper.%
%%\todo{Reveal a bit about this paper?}
%
%%\begin{remark} \label{re:type-signature}
%%Most of existing theories treat types independently of terms, thereby excluding them from signatures.
%%To the best of our knowledge, the only exception to this approach is found in the work of~\citet{Arkor2020}, which incorporates signatures for both terms and types.
%%Interestingly, this inclusion is also critical for type synthesis for comparing a concrete type $N \bto N$ with an abstract type $A \bto B$, where $A$ and $B$ are type variables in a typing rule.
%%\end{remark}

>>>>

\section{Discussion} \label{sec:future}
We believe that our formal treatment lays a foundation for further investigation, as the essential aspects of bidirectional typing have been studied rigorously. 
While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and extensions remain unexplored.
We will discuss related work and outline potential directions for future research.
\subsubsection{Language formalisation frameworks}
Many general definitions of logics and type theories and frameworks for defining them have been proposed.

\citet{Harper1993a,Harper2007}
\cite{Uemura2021}
\cite{Bauer2020,Haselwarter2021,Bauer2022a}
Our general definitions are akin to the definitions used in other language formalisation frameworks~\citep{Ahrens2022,Allais2021,Fiore2022} in a proof assistant---the concept of a signature is at object level.

%\subsubsection{Generic bidirectional type synthesis beyond simple types}
%Clearly, bidirectional type synthesis becomes prominent because it can scale up to deal with more complex types than simple types.
%As for a general treatment for another class of languages, one has to devise a general definition of languages, which may have been used by existing language-formalisation frameworks.
%Then, one could enrich the general definition with modes and adopt the definition of mode-correctness accordingly. 
%Soundness and completeness of our formulation should carry over, since they are only about the separation and combination of mode and typing information for a given raw term (in a syntax-directed formulation).
%Mode decoration is about annotating a raw term with modes and marking missing annotations, used for guiding the direction during bidirectional type synthesis, so it should also carry over.
%What makes the general development of bidirectional type synthesis challenging is inherent to the tension between expressiveness and usability for type synthesis and the general treatment for other language facilities.
%We discusses the cases for dependent types and for polymorphic types below.
%
%\paragraph{Dependent types} Type synthesis for a dependent type theory involves term equality check for type equality problem, so a generic decidable bidirectional type synthesis relies on the decidability of the term equality check which in turn requires the ability of specifying computation rules within a language specification.
%\varcitet{Felicissimo2023}{'s computational LF} is capable of defining dependent type theories with computational rules like \textsc{Dedukti}~\cite{Assaf2016} and generic bidirectional type synthesis.
%Apart from their more general definition targeting dependent type theories, generic bidirectional type synthesis \cite[Theorem~3.12]{Felicissimo2023} that is decidable, sound, and complete (with respect to decorated terms) is possible under the assumption that a given language specification is mode-correct (or well-typed \cite[Section~3.2]{Felicissimo2023}) and equality check is well-behaved: the specified computation rules are type-preserving, confluent, and strongly normalising.
%Yet, the last assumption is hard to establish even for a specific language and still lacks a general consideration.
%
%\paragraph{Polymorphic types} For \SystemF and other languages where type-level variable binding is allowed, we can start with the notion of polymorphic signature $(\Sigma, \Omega)$ by~\citet{Hamana2011} whereas the arity of each type construct $\sigma$ in $\Sigma$ becomes a binding arity with a single sort $*$ and accordingly a term construct can specify a pair of extension contexts~$\left<\Xi\right>[\Delta]$ for term variables $\Delta$ and type variables $\Xi$.
%It is not hard to see how general definitions for bidirectional typing rules and mode derivations can be extended from \varcitet[Section~3.4]{Hamana2011}{'s definitions}.
%For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can then be specified as operations
%\[
%  \mathsf{all}\colon *\rhd [*]* \to *
%  \quad\text{and}\quad
%  \mathsf{tabs}\colon [*]A \rhd \left< * \right>A^{\chk} \to
%  \tyOp_{\mathsf{all}}(\alpha.\, A)^{\chk}
%\]
%\Cref{thm:bidirectional-type-synthesis-checking} should also carry over, since no equation is imposed on types and no guess work is needed.
%However, explicit type application is unusable in practice but implicit application results the \emph{instantiation problem}~\cite[Section~5]{Dunfield2021} which amounts to replacing type equality check with checking the subtyping relation $A <: B$ in rule $\ChkRule{Sub}$ and type synthesis with the guess of $B$ in $\forall \alpha.\, A <: A[B/\alpha]$. 
%A satisfactory theory for bidirectional typing with polymorphism and subtyping should be neutral and thus accommodate various solutions to the instantiation problem, which is still under development.
%
\subsubsection{Generic bidirectional type synthesis beyond simple types}

Bidirectional type synthesis plays a crucial role in handling more complex types than simple types.
To provide a formal treatment for a broader class of languages, we need a general definition of languages in question, which may have been used in existing theories or frameworks.
Then, this general definition can be augmented with modes and the concept of mode-correctness can be introduced.
Soundness and completeness (\cref{lem:soundness-completeness}) should naturally extend, as they pertain solely to the separation and combination of mode and typing information within a given raw term in a syntax-directed formulation.
Mode decoration (\cref{sec:mode-decoration}) involves annotating a raw term with modes and marking missing annotations should also work.
The challenge in developing bidirectional type synthesis on a more general case comes from striking the right balance between expressiveness and usability for type synthesis while considering other language features.
We discuss cases involving dependent types and polymorphic types below.

\paragraph{Dependent Types}
In the context of dependent type theory, type synthesis entails a term equality check for type equality.
Achieving generic decidable bidirectional type synthesis depends on the decidability of term equality, which, in turn, relies on the ability to specify computation rules within a language specification.

\varcitet{Felicissimo2023}{'s computation LF} is capable of defining dependent type theories with computational rules, similar to Dedukti, and enabling generic bidirectional type synthesis.
Generic bidirectional type synthesis, as presented in Felicissimo's work, is decidable, sound, and complete with respect to decorated terms, provided that the underlying language specification is mode-correct, and the equality check is well-behaved, i.e.\ type-preserving, confluent, and strongly normalising.
However, the last assumption is challenging to establish, even for specific languages and lacks a comprehensive general understanding.

\paragraph{Polymorphic Types}
In the case of languages like \SystemF and others that permit type-level variable binding, we can start with the notion of a polymorphic signature, as introduced by Hamana---each type construct in a signature is specified by a binding arity with a single sort $*$ and a term construct can employ a pair of extension contexts for term variables and type variables.

Extending general definitions for bidirectional typing and mode derivations from \citeauthor{Hamana2011}'s work is straightforward. 
For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations:
\[
\mathsf{all} : * \rhd [*] * \to *
\quad\text{and}\quad
\mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha. A)^{\chk}
\]
The decidability of bidirectional type synthesis checking (\cref{thm:bidirectional-type-synthesis-checking}) should also carry over, as no equations are imposed on types and no guessing is required.

However, explicit type application is impractical but implicit application results in the \emph{instantiation problem} which amounts to replacing type equality checks $A = B$ with checking subtyping $A <: B$ and guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$ in the presence of universal quantification.
Developing a general theory for bidirectional typing with polymorphism and subtyping that accommodates various solutions to the instantiation problem is still under development.

\subsubsection{More characteristics of bidirectional typing}

\subsubsection{Conclusion}
There are still many concepts that are possible to articulate such as, for example, Pfenning's recipe for bidirectionalising typing rules.
In contrast, there are concepts that may be hard to pin down, notably `annotation character'~\cite{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs.
These considerations may continue to require creativity from the designer of type systems whereas its implementation could be automated entirely.
